Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-10-11 manu-draft.Rmd
Address some abbreviations.
Edited some instances of abbreviations. The Environmental Protection Commission of Hillsborough County typically prefers EPCHC, as their abbreviation. But, I didn't modify from what you have included.

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer used in food production [@Henderson04].  By 2001, an estimated 40 million tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.  The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack with three holding ponds located less than two miles from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].  Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, HRK Holdings, LLC, with oversight by the Florida Departmental of Environmental Protection (FDEP).  Decreasing holding capacity of the ponds from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee have contributed to degradation of the facility.  Discharges of wastewater from the stacks occurred in the early 2000s and 2011 to nearby Bishop Harbor connected to Tampa Bay.  Those discharges resulted in spatially-restricted, ecosystem responses [@Switzer11;@Garrett11].

@@

HRK Holdings, LLC (hereafter, HRK) acquired Piney Point in August 2006 through an administrative agreement with FDEP.  This agreement transferred responsibility of the site to HRK with the intention that any future uses must protect and be compatible with the integrity of stack closure and long-term care.  In 2011, HRK agreed to the storage of 1.5 million cubic yards of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).  This material was added to an existing gypstack at Piney Point. Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency discharge that released 169 million gallons of dredged saltwater slurry and 3.5 tons of nitrogen to receiving waters leading to Bishop Harbor. The dredging and deposit of slurry at Piney Point continued following structural fortifications to the holding stacks to ensure integrity with additional loadings.  HRK maintains ownership and responsibility of the site to present day with oversight by FDEP.

@@

Water quality data in Tampa Bay have been collected at fixed sampling sites since 1974 by the Environmental Protection Commission of Hillsborough County.  These data include samples at 45 stations used by the TBEP to assess annual progress towards programmatic goals and regulatory thresholds applicable for each of the major segments in Tampa Bay.  Discrete water samples are collected monthly at mid-depth and processed in the laboratory immediately after collection. Similarly, the Parks and Natural Resources Department of Manatee County have been collecting data in Tampa Bay at locations south of Piney Point and south of the main dredged channel of Tampa Bay (approximate longitudinal axis of the bay).  The sampling design is similar to the EPC data, with the exception that sites are sampled approximately every three months but at a higher spatial density per unit area.  Additionally, the period of record for monitoring data from Manatee County began in 1996.  Long-term monitoring stations for Hillsborough and Manatee Counties are shown in Figure \@ref(fig:map)a.

>>>>

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer used in food production [@Henderson04].  By 2001, an estimated 40 million tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.  The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack with three holding ponds located less than two miles from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].  Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, HRK Holdings, LLC (hereafter HRK), with oversight by the Florida Departmental of Environmental Protection (FDEP).  Decreasing holding capacity of the ponds from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee have contributed to degradation of the facility.  Discharges of wastewater from the stacks occurred in the early 2000s and 2011 to nearby Bishop Harbor connected to Tampa Bay.  Those discharges resulted in spatially-restricted, ecosystem responses [@Switzer11;@Garrett11].

@@

HRK acquired Piney Point in August 2006 through an administrative agreement with FDEP.  This agreement transferred responsibility of the site to HRK with the intention that any future uses must protect and be compatible with the integrity of stack closure and long-term care.  In 2011, HRK agreed to the storage of 1.5 million cubic yards of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).  This material was added to an existing gypstack at Piney Point. Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency discharge that released 169 million gallons of dredged saltwater slurry and 3.5 tons of nitrogen to receiving waters leading to Bishop Harbor. The dredging and deposit of slurry at Piney Point continued following structural fortifications to the holding stacks to ensure integrity with additional loadings.  HRK maintains ownership and responsibility of the site to present day with oversight by FDEP.

@@

Water quality data in Tampa Bay have been collected at fixed sampling sites since 1974 by the EPC.  These data include samples at 45 stations used by the TBEP to assess annual progress towards programmatic goals and regulatory thresholds applicable for each of the major segments in Tampa Bay.  Discrete water samples are collected monthly at mid-depth and processed in the laboratory immediately after collection. Similarly, the Parks and Natural Resources Department of Manatee County have been collecting data in Tampa Bay at locations south of Piney Point and south of the main dredged channel of Tampa Bay (approximate longitudinal axis of the bay).  The sampling design is similar to the EPC data, with the exception that sites are sampled approximately every three months but at a higher spatial density per unit area.  Additionally, the period of record for monitoring data from Manatee County began in 1996.  Long-term monitoring stations for Hillsborough and Manatee Counties are shown in Figure \@ref(fig:map)a.
